---
title: "Lab 7: Group 03"
format: html
editor: visual
---

### Group members:

s225049: Ida Sofie Goldschmidt\
s225074: Annekatrine Kirketerp-Møller\
s225051: Emil August Dissing Berntsen\
s224974: Rasmus Kamp\
s215109: Rolf Larsen

## Load libraries

```{r}
#| message: false

library("tidyverse")
library("broom")
library("cowplot")

```

## Loading and tidying data:

```{r}
#| message: false

biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")
```

To do the PCA, we first need to remove all non-numeric columns, since the function *prcomp* does not take non-numeric inputs. The PCA is then stored in the function *pca_fit*. We also need to scale the data, to make sure that the difference in value ranges does not interfere with the analysis.

```{r}
pca_fit <- biopsy |> 
  select(where(is.numeric)) |> # retain only numeric columns
  prcomp(scale = TRUE) # do PCA on scaled data
```

```{r}
pca_fit |>
  augment(biopsy) |> # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = outcome)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(malignant = "#D55E00", benign = "#0072B2")
  ) +
  theme_half_open(12) + background_grid()
```

```{r}
# extract rotation matrix
pca_fit |>
  tidy(matrix = "rotation")
```

```{r}
# define arrow style for plotting
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# plot rotation matrix
pca_fit |>
  tidy(matrix = "rotation") |>
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") |>
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "#904C2F"
  ) +
  xlim(-1.25, .5) + ylim(-.5, 1) +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal_grid(12)
```

```{r}
pca_fit |>
  tidy(matrix = "eigenvalues")
```

```{r}
pca_fit |>
  tidy(matrix = "eigenvalues") |>
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#56B4E9", alpha = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  theme_minimal_hgrid(12)
```
